Type theory in type theory with single substitutions

### 2025-01-30 cpp/p.tex
stuff

Some advantages of using a GAT presentation:
\begin{itemize}
\item  In type theory, typing depends on conversion:
  in an extrinsic presentation, this is expressed by a rule such as
  the following.
  \[
  \infer{\Gamma\vdash t : B}{A \sim B && \Gamma\vdash t : A}
  \]
  When conversion is equality, this rule holds by the general
  properties of equality; and we also avoid lots of boilerplate
  stating that e.g.\ conversion is a congruence with respect to all
  operations (for an intrinsic presentation with explicit conversion
  relation see \cite{chapman09eatitself}).
\item Syntax and semantics are concordant: the syntax is simply the
  initial model, which always exists for any GAT and is called a
  quotient inductive-inductive type
  \cite{DBLP:journals/pacmpl/KaposiKA19}. The fact that the syntax is
  quotiented means that any function defined on it needs to respect
  the equations. For example, it is not possible to define a printing
  function which distinguishes the terms $(\lambda x.x)\,1$ and $1$,
  as these are convertible. However, it is still possible to define
  normalisation \cite{lmcs:4005} and typechecking \cite{typecheck} at
  this level of abstraction.
\item The GAT approach is also more abstract than extrinsic
  approaches, which means that we don't have to make ad-hoc
  choices. For example, we do not have to decide whether we want
  function space {\`a} la Curry or {\`a} la Church, whether we work
  with paranoid or economic typing rules \cite[Section
    5]{winterhalter:hal-01626651}, or whether we have explicit or
  recursively defined substitution (instantiation of variables by
  terms, written $t[x\mapsto u]$ or $t[u/x]$).  These choices are
  forced by the notion of GAT: the domain of a lambda appears in its
  type, se we are always {\`a} la Church; algebraic theories don't
  allow recursively defined operators, so instantiation is an explicit
  operation, and so on.
\end{itemize}

@@

However, even at the GAT level of abstraction, there are some choices
to be made about the instantiation operation. It is usually convenient
to use parallel substitutions, which means that the instantiation
operation replaces all variables in a context by terms in another
context \emph{at the same time}, that is, $t[x_0\mapsto
  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are all the free
variables in $t$. $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$ is called a
parallel substitution, and the terms $u_0,\dots,u_n$ are all in the
same context. Parallel substitutions are very natural in the algebraic
setting because they form the morphisms in a category where objects
are contexts (lists of types), and then it is clear what equations
they need to satisfy. Some examples of algebraic definitions of type
theory using parallel substitutions ranging from the more syntactic to
the more semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard},
categories with families (CwF
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), Awodey's natural
models \cite{DBLP:journals/mscs/Awodey18}, contextual categories
\cite{DBLP:journals/apal/Cartmell86}, C-systems
\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
\cite{10.1145/3204492}.

In traditional extrinsic presentations
(e.g.\ \cite{alti:phd93,DBLP:books/daglib/0005958}) instantiation
simply replaces all free occurrences of a variable $x$ with a term $u$
as in the notation $t[x\mapsto u]$. A substitutions consists of a
variable name (a pointer into a context) and a term. Substitutions
cannot be composed, thus they do not form a category. To bridge the
gap between the intrinsic parallel substitution world and the
extrinsic single substitution world, Voevodsky introduced B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
desription of a single substitution calculus. B-systems involve
complex rules describing telescopes, weakenings and substitutions
under telescopes, with several equations.

In this paper, we introduce a new single substitution calculus (SSC)
which is simpler and more minimalistic than the existing algebraic
approaches. The instantiation operation $\blank[\blank]$ takes either a
single substitution or a single weakening as an argument, and there
are eight equations explaining how these behave: four of these say how
to instantiate variables and the other four equations are needed to
typecheck the first four. The SSC shows that to explain type theory in
an algebraic way, there is no need for categories, parallel
substitutions or weakenings, empty substitution, telescopes, or
combinations of these. All our operations are easy to motivate, we
showcase this in Section \ref{sec:tt} which is a tutorial introduction
to our theory. We believe that our SSC is in a sweet spot: our
notations are close to traditional extrinsic notations, but they are
algebraic, thus come with a well behaved model theory, and can be
easily related to more semantic descriptions.

Our SSC is in some sense too minimalistic: for example, the equation
$b[\p^+][\langle\q\rangle] = b$ is not \emph{derivable} in any model,
but it is \emph{admissible} in the syntax. (Here $b : \Tm\,(\Gamma\ext
A)\,B$ is a term which depends on a context $\Gamma$ and an extra
variable of type $A$; $b[\p^+] : \Tm\,(\Gamma\ext A\ext
A[\p])\,(B[\p^+])$ is a version where we weakened \emph{before} the
last variable; in $b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$
we substituted the last variable for the previous one.)  This is
analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In this paper, we show
that the \emph{syntax} of our single substitution theory and the
syntax of the corresponding parallel substitution theory (CwF-based
theory) are isomorphic. All CwF-based models are models of our single
substitution theory, but not necessarily the other way. While
B-systems are equivalent to CwFs (and C-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more SSC models
than CwF-based models. This relationship is similar to that of
extensional combinatory calculus and lambda calculus, where the
syntaxes are equivalent, and every lambda model is a combinatory
model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. Another example is the
relationship between monoids over a set $X$ and nil/cons algebras over
$X$, where every monoid is a nil/cons-algebra, but not the other way;
but it is well-known that free monoids over $X$ (syntax of monoid over
$X$) are the same as lists of $X$ (the syntax of nil/cons algebras
over $X$).

@@

Our SSC has 8 equations, and these are enough to describe type
theories with arbitrary choice of type formers (see the description of
SOGATs in Subsection \ref{sec:related}). In the particular case of
type theory with a Coquand-universe and $\Pi$ types, we can do better:
4 (conditional) equations on variables are enough, because using the
fact that every type is represented by a terma, we can derive the other
4 equations.

If in addition to having a universe and $\Pi$ types, our type theory
also supports unit and $\Sigma$ types, we can actually derive a
CwF-based model from an SSC-model. The idea is that $\Sigma$ types can
represent contexts, functions between $\Sigma$ types represent
parallel substitutions, functions into the universe represent depenent
types. Then the SSC is just there to formulate $\Sigma$, $\Pi$, $\U$,
and then we use these to define a parallel substitution calculus. So
although SSC models and CwFs without any type formers are not
equivalent, if we equip them with enough type formers, they become
(weakly) equivalent.

All of our constructions can be understood as happening internal to
extenional type theory, and most results in this paper are formalised
in Agda. Most results in this paper were presented at the TYPES 2024
conference \cite{singleTypes}.

\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus. We show that
  there is no need for parallel substitutions, empty substitution,
  parallel weakenings, telescopes, or combinations of these to
  describe the syntax of type theory in an intrinsic way. We present
  our calculus in an easy-to-understand way.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item For type theory with $\Pi$, $\U$, a minimised presentation of
  the equations which results in an isomorphic theory.
\item For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
  single and parallel models are equivalent in a weak sense.
\end{itemize}

@@

After discussing related work, in Section \ref{sec:tt}, we introduce
our single substitution calculus (SSC) only relying on a working
knowledge of an implementation of type theory such as Agda, no prior
experience with the metatheory of type theory is required. Section
\ref{sec:admissible} shows that in the syntax of SSC, several new
equations are admissible, and that the syntax of SSC is actually
isomorphic to the CwF-based syntax. In Section \ref{sec:minimisation}
we minimise our SSC obtaining a theory with fewer equations, relying
on the presence of $\Pi$ and $\U$ in our theory. In Section
\ref{sec:cwf}, we show that in the presence of $\Pi$, $\U$ and
$\Sigma$, SSC-models are actually equivalent to CwF-models in a weak
sense. We conclude in Section \ref{sec:conclusion}.

\subsection{Related work}

@@

substitutions. B-systems correspond to our SSC as essentially
algebraic theories correspond to generalised algebraic theories, or
sets with a map into $I$ correspond to indexed families over $I$

@@

order. However we don't have the lifted versions of our equations,

@@

\paragraph{Ehrhard's calculus}
The first generalised algebraic presentation of type theory was Thomas
Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
parallel substitution calculus with $\p$, $\blank^+$ and
$\langle\blank\rangle$ operations, just like in our calculus. Our
single substitution calculus is essentially Ehrhard's calculus with
the categorical composition and identity operations
removed. Categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
$\blank,\blank$ operations which make it more apparent that
substitutions are lists of terms. CwFs are equivalent to Ehrhard's
calculus, and so are contextual categories/C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.

\paragraph{Formalisations of type theory}
Most computer formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than ordinary Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. Although there is no general proof
published about this, we have experimental evidence that QIITs are
supported by the cubical set model \cite{DBLP:conf/lics/CoquandHM18}
and the setoid model \cite{kaposi-qiit-setoid}, thus Cubical Agda
\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
theory have been formalised \cite{cohtt}. Without QIITs, tricks such
as shallow embedding the syntax can be used to formalise metatheoretic
results about type theory such as gluing
\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

\paragraph{SOGATs}
We can define languages with binders as second-order theories
following higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99}
and logical frameworks
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}. Second-order
generalised algebraic theories (SOGATs
\cite{uemura,DBLP:conf/fscd/KaposiX24}) can be used to present type
theory even more abstractly than GATs: they even abstract over the way
of defining the substitution calculus: contexts, substitutions,
instantiations are simply modelled by the metatheoretic function
space. For example, a type theory with $\Pi$, Coquand-universes and a
universe level lifting operation is described by the following SOGAT
(where $\cong$ means isomorphism):
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
% Adding $\top$ and $\Sigma$ types (Definition \ref{def:sigma}) is
% described as the following extension of the above SOGAT:
% \begin{equation}\label{eq:sigma}
%   \begin{alignedat}{10}
%     & \top && : \Ty\,0 \\
%     & \tt && : \Tm\,\top \\
%     & \top\eta && : t = \tt \\
%     & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
%     & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
% \end{alignedat}
% \end{equation}
Second-order models of SOGATs are not well-behaved as there is no good
notion of homomorphism between them. This is why Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: each GAT
comes with a category of models with initial model (syntax), (co)free
models, and so on. Actually, Kaposi and Xie define two different
translations: one based on parallel substitutions and one based on
single substitutions. If we apply their parallel translation to the
SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set
\end{alignedat}
\end{equation}
with only two sorts and no operations or equations, we obtain the GAT
of category with families (CwF). If we use their single substitution
translation on the SOGAT (\ref{eq:tt}), we obtain Definition
\ref{def:tt} (also listed in Appendix \ref{app:tt}). This paper can be
seen as taking a big enough type theory (described as a SOGAT), and
investigating how the result of its single substitution translation
relates to the result of its parallel translation (the latter
first-order theory is quite well-known).

There are techniques to prove properties of type theories without
choosing between parallel or single substitutions, and staying at the
SOGAT level of abstraction: Synthethic Tait Computability
\cite{DBLP:phd/us/Sterling22} or internal sconing
\cite{DBLP:conf/fscd/BocquetKS23}.

>>>>

Some advantages of using a GAT presentation:
\begin{itemize}
\item  In type theory, typing depends on conversion:
  in an extrinsic presentation, this is expressed by a rule such as
  the following.
  \[
  \infer{\Gamma\vdash t : B}{\Gamma\vdash A \sim B && \Gamma\vdash t : A}
  \]
  When conversion $\sim$ is equality, this rule holds by the general
  properties of equality; and we also avoid lots of boilerplate
  stating that e.g.\ conversion is a congruence with respect to all
  operations (for an intrinsic presentation with explicit conversion
  relation see \cite{chapman09eatitself}).
\item Syntax and semantics are concordant: the syntax is simply the
  initial model, which always exists for any GAT and is called a
  quotient inductive-inductive type
  \cite{DBLP:journals/pacmpl/KaposiKA19}. The fact that the syntax is
  quotiented means that any function defined on it needs to respect
  the equations. For example, it is not possible to define a printing
  function which distinguishes the terms $(\lambda x.x)\,1$ and $1$,
  as these are convertible. However, it is still possible to define
  normalisation \cite{lmcs:4005} and typechecking \cite{typecheck} at
  this level of abstraction.
\item The GAT approach is also more abstract than extrinsic
  approaches, which means that we don't have to make ad-hoc
  choices. For example, we do not have to decide whether we want
  function space {\`a} la Curry or {\`a} la Church; whether we work
  with paranoid or economic typing rules \cite[Section
    5]{winterhalter:hal-01626651}; whether we have explicit or
  recursively defined substitution (instantiation of variables by
  terms, written $t[x\mapsto u]$ or $t[u/x]$).  These choices are
  forced by the notion of GAT: the domain of a lambda appears in its
  type, so we are always {\`a} la Church; algebraic theories don't
  allow recursively defined operators, so instantiation is an explicit
  operation; and so on.
\end{itemize}

@@

However, even at the GAT level of abstraction, there are some choices
to be made about the instantiation operation. It is usually convenient
to use parallel substitutions, which means that the instantiation
operation replaces all variables in a context by terms in another
context \emph{at the same time}, that is, $t[x_0\mapsto
  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are the variables
in the context of $t$. $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$ is
called a parallel substitution, where the terms $u_0,\dots,u_n$ are
all in the same context. Parallel substitutions are very natural in
the algebraic setting because they form the morphisms in a category
where objects are contexts (lists of types), then types/terms are
preheaves, context extension is a representability condition, so the
equations of the algebraic theory are forced by the categorical
structure. Some examples of algebraic definitions of type theory using
parallel substitutions ranging from the more syntactic to the more
semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories
with families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}),
Awodey's natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
\cite{10.1145/3204492}.

In traditional extrinsic presentations
(e.g.\ \cite{alti:phd93,DBLP:books/daglib/0005958}) instantiation
replaces free occurrences of a single variable $x$ with a term $u$ as
in the notation $t[x\mapsto u]$. A substitution consists of a
variable name (a pointer into a context) and a term. Substitutions
cannot be composed, thus they do not form a category. To bridge the
gap between the intrinsic parallel substitution world and the
extrinsic single substitution world, Voevodsky introduced B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
desription of a single substitution calculus. B-systems involve
complex rules describing telescopes, weakenings and substitutions
under telescopes, with several equations.

In this paper, we introduce a new single substitution calculus (SSC)
which is simpler and more minimalistic than existing algebraic
approaches. The instantiation operation $\blank[\blank]$ takes either a
single substitution or a single weakening as an argument, and there
are eight equations explaining how these behave: four of these say how
to instantiate variables and the other four equations are needed to
typecheck the first four. The SSC shows that to explain type theory in
an algebraic way, there is no need for categories, parallel
substitutions or weakenings, empty substitution, telescopes, or
combinations of these. All our operations are easy to motivate, we
showcase this in Section \ref{sec:tt} which is a tutorial introduction
to our theory. We believe that our SSC is in a sweet spot: our
notations are close to traditional extrinsic notations, but they are
algebraic, thus come with a well behaved model theory, and can be
easily related to more semantic descriptions.

Our SSC is in some sense too minimalistic: for example, the equation
$b[\p^+][\langle\q\rangle] = b$ is not \emph{derivable} in any model,
but it is \emph{admissible} in the syntax. (Here $b : \Tm\,(\Gamma\ext
A)\,B$ is a term which depends on a context $\Gamma$ and an extra
variable of type $A$; $b[\p^+] : \Tm\,(\Gamma\ext A\ext
A[\p])\,(B[\p^+])$ is a version where we weakened \emph{before} the
last variable; in $b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$
we substituted the last variable for the previous one.)  This is
analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In this paper, we show
that the \emph{syntaxes} of our SSC and the corresponding parallel
substitution calculus (CwF-based calculus) are isomorphic. All
CwF-based models are models of our SSC, but not
necessarily the other way. While B-systems are equivalent to CwFs (and
C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
SSC models than CwF-based models. This relationship is similar to that
of extensional combinatory calculus and lambda calculus, where the
syntaxes are equivalent, and every lambda model is a combinatory
model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. Another example is the
relationship between monoids over a set $X$ and nil/cons algebras over
$X$, where every monoid is a nil/cons-algebra, but not the other way;
but it is well-known that free monoids over $X$ (syntax of monoid over
$X$) are the same as lists of $X$ (the syntax of nil/cons algebras
over $X$).

@@

Our SSC has 8 equations, and these are enough to describe type
theories with arbitrary choice of type formers (see the paragraph on
SOGATs in Subsection \ref{sec:related}). In the particular case of
type theory with a Coquand-universe and $\Pi$ types, we can do better:
the 4 (conditional) equations are enough, because we can derive the
other 4 equations using the fact that every type is represented by a
term.

If in addition to having a universe and $\Pi$ types, our type theory
also supports unit and $\Sigma$ types, we can actually derive a
CwF-based model from an SSC-model. The idea is that $\Sigma$ types can
represent contexts, functions between $\Sigma$ types represent
parallel substitutions, functions into the universe represent
dependent types. Then the SSC is just there for bootstrapping
purposes: we use the SSC to formulate $\Sigma$, $\Pi$, $\U$, and then
we use these to define a parallel substitution calculus. In summary,
although SSC models and CwFs without any type formers are not
equivalent, if we equip them with enough type formers, they become
(weakly) equivalent.

All of the constructions in this paper can be understood as happening
internal to extensional type theory, and most results were formalised
in Agda. Some results in this paper were presented at the TYPES 2024
conference \cite{singleTypes}.

In summary, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus. Our calculus
  does not feature parallel substitutions/weakenings, empty
  substitution, telescopes. We present our calculus in an
  easy-to-understand way where all operations are well-motivated.
\item The $\alpha$-normalisation technique which shows that the syntax
  of our single substitution calculus is isomorphic to the syntax of
  the CwF-based theory.
\item For type theory with $\Pi$, $\U$, a minimised presentation of
  the equations which results in an isomorphic theory.
\item For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
  single and parallel models are equivalent in a weak sense.
\end{itemize}

@@

After discussing related work and our metatheory, in Section
\ref{sec:tt}, we introduce our single substitution calculus (SSC) only
relying on a working knowledge of an implementation of type theory
such as Agda, no prior experience with the metatheory of type theory
is required. Section \ref{sec:admissible} shows that in the syntax of
SSC, several new equations are admissible, and that the syntax of SSC
is actually isomorphic to the CwF-based syntax. In Section
\ref{sec:minimisation}, we minimise our SSC obtaining a theory with
fewer equations, relying on the presence of $\Pi$ and $\U$ in our
theory. In Section \ref{sec:cwf}, we show that in the presence of
$\Pi$, $\U$, $\Sigma$, SSC-models are actually equivalent to
CwF-models in a weak sense. We conclude in Section
\ref{sec:conclusion}.

\subsection{Related work}\label{sec:related}

@@

substitutions. B-systems relate to our SSC as essentially
algebraic theories relate to generalised algebraic theories, or
sets with a map into $I$ relate to indexed families over $I$

@@

order. However we don't include the lifted versions of these equations,

@@

\paragraph{Ehrhard's calculus}
The first generalised algebraic presentation of type theory was Thomas
Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
parallel substitution calculus with $\p$, $\blank^+$ and
$\langle\blank\rangle$ operations, just like our calculus. Our
single substitution calculus is essentially Ehrhard's calculus with
the categorical composition and identity operations
removed. Categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
$\blank,\blank$ operations which make it more apparent that
substitutions are lists of terms. CwFs are equivalent to Ehrhard's
calculus, and so are contextual categories/C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.

\paragraph{Formalisations of type theory}
Most computer formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than plain Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. Although we don't have a general
proof, we have experimental evidence that QIITs are supported by the
cubical set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid
model \cite{kaposi-qiit-setoid}, thus Cubical Agda
\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
theory have been formalised \cite{cohtt}. Without QIITs, tricks such
as shallow embedding the syntax can be used to formalise metatheoretic
results about type theory such as gluing
\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

\paragraph{SOGATs}
We can define languages with binders as second-order theories
following higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99}
and logical frameworks
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}. Second-order
generalised algebraic theories (SOGATs
\cite{uemura,DBLP:conf/fscd/KaposiX24}) can be used to present type
theory more abstractly than GATs: SOGATs abstract over the exact
definition of the substitution calculus (whether substitutions are
single or parallel; whether we have CwF-style or Ehrhard-style
operations): contexts, substitutions, instantiations are simply
modelled by the metatheoretic function space. For example, a type
theory with $\Pi$, Coquand-universes and a universe level lifting
operation is described by the following SOGAT (where $\cong$ means
isomorphism):
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
% Adding $\top$ and $\Sigma$ types (Definition \ref{def:sigma}) is
% described as the following extension of the above SOGAT:
% \begin{equation}\label{eq:sigma}
%   \begin{alignedat}{10}
%     & \top && : \Ty\,0 \\
%     & \tt && : \Tm\,\top \\
%     & \top\eta && : t = \tt \\
%     & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
%     & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
% \end{alignedat}
% \end{equation}
Second-order models of SOGATs are not well-behaved, as there is no
good notion of homomorphism between them. This is why Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: a model of a
SOGAT then is a model of the GAT the we obtain by their
translation. GATs come with well-behaved metatheory: a category of
models witha initial model (syntax), (co)free models
\cite{andras,DBLP:phd/hal/Moeneclaey22}, and so on. However, the
translation is not unique: Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} define two different translations, one
based on parallel substitutions and one based on single
substitutions. If we apply their parallel translation to the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set
\end{alignedat}
\end{equation}
with only two sorts and no operations or equations, we obtain the GAT
of category with families (CwF). If we use their single substitution
translation to the SOGAT (\ref{eq:tt}), we obtain Definition
\ref{def:tt} (also listed in Appendix \ref{app:tt}). The current paper
can be seen as taking a big enough type theory (described as a SOGAT),
and investigating how the result of its single substitution
translation relates to the result of its parallel translation (the
latter first-order theory is quite well-known). There is no general
result about the relationship of the two translations.

There are techniques to prove properties of type theories without
choosing between parallel or single substitutions, and staying at the
SOGAT level of abstraction: Synthethic Tait Computability
\cite{DBLP:phd/us/Sterling22} and internal sconing
\cite{DBLP:conf/fscd/BocquetKS23}.
